Path Analysis Data Development - Chapter 2

### 2016-03-14 sem-data-development.Rmd
Add lagged temperature and rainfall variables.

# Untagged dbh data
untagged_dbh_2005 <- read.csv("analysis/data/untagged_dbh2005_summary.csv")
untagged_dbh_2014 <- read.csv("analysis/data/untagged_dbh2014_summary.csv")
canopy_cover <- read.csv("analysis/data/plot_canopy_cover_dens.csv")
```
```{r load physical data, echo=FALSE, message=F}
# Physical environment: topography
library(rgdal); library(plyr); library(dplyr)
plots_shp <- readOGR("analysis/data/plot_gis/", "soco_plots_metric")
plots_env <- plots_shp@data
summary(plots_env)
plots_env <- select(plots_env, -cat, -X, -Y)

# Elevation and vegetation height are in feet, slope is degrees, and canopy density is the proportion of the cell covered by canopy (0-1)
plots_topo.shp <- readOGR("analysis/data/plot_gis/", "plots202_vegtopo")
plots_vegtopo <- plots_topo.shp@data
summary(plots_vegtopo)
plots_vegtopo <- select(plots_vegtopo, -cat) %>% rename(lon_utm = X, lat_utm = Y)
```
```{r load weather data, echo=FALSE}
# Temperature: rainy season hour summaries (long format for each year)
rs_temps <- read.csv("analysis/data/rs_select_temps.csv")
summary(rs_temps)
rs_temps <- select(rs_temps, -X)

summer_temps <- read.csv("analysis/data/summer_select_temps.csv")
summary(summer_temps)
summer_temps <- select(summer_temps, -X)

# Rainfall: interpolated/estimated rainfall for each season
load("~/Documents/soco_ppt/plot_ppt_estimates.RData")
summary(ppt_days)
summary(ppt_total)
# plots_rain <- read.csv("analysis/data/plot202rain.csv")
# summary(plots_rain)
# plots_rain <- select(plots_rain, -cat)
# pairs(select(plots_rain, -PLOT_ID), 
#       lower.panel = panel.cor, 
#       diag.panel = panel.hist)
ppt_days@data <- rename(ppt_days@data, plotid = PLOT_ID)
ppt_total@data <- rename(ppt_total@data, plotid = PLOT_ID)
```

## Join remotely sensed and derived physical data
LiDAR products, variables derived from DEMs
```{r join remotely sensed data}
plots_env <- inner_join(plots_env, plots_vegtopo, by = "PLOT_ID")
summary(plots_env)
plots_env$elev_m <- plots_env$elev15m * .3048
# plots_env <- select(plots_env, -elevation, -tmi) #remove elevation, tmi from old derivation (I think a legacy 10m USGS DEM)
plots_env$veght_m <- plots_env$veght15m * .3048
rm(plots_vegtopo)
```

## Convert rain data from wide to long
```{r rain wide to long}
library(tidyr)
rain2d_total <- ppt_total@data %>% select(plotid, ends_with("2d"))
names(rain2d_total) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain2d_total <- rain2d_total %>% gather(year, rain_tot_2d, -plotid)

rain3d_total <- ppt_total@data %>% select(plotid, ends_with("3d"))
names(rain3d_total) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain3d_total <- rain3d_total %>% gather(year, rain_tot_3d, -plotid)

rain_v_total <- ppt_total@data %>% select(plotid, ends_with("v"))
names(rain_v_total) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain_v_total <- rain_v_total %>% gather(year, rain_tot_v, -plotid)

rain_r_total <- ppt_total@data %>% select(plotid, ends_with("r"))
names(rain_r_total) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain_r_total <- rain_r_total %>% gather(year, rain_tot_r, -plotid)

rain_psm_total <- ppt_total@data %>% select(plotid, contains("sum"))
names(rain_psm_total) <- c("plotid", paste(c(2004:2010), sep = ","))
rain_psm_total <- rain_psm_total %>% gather(year, rain_tot_psm, -plotid)

rain2d_days <- ppt_days@data %>% select(plotid, ends_with("2d"))
names(rain2d_days) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain2d_days <- rain2d_days %>% gather(year, rain_days_2d, -plotid)

rain3d_days <- ppt_days@data %>% select(plotid, ends_with("3d"))
names(rain3d_days) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain3d_days <- rain3d_days %>% gather(year, rain_days_3d, -plotid)

rain_v_days <- ppt_days@data %>% select(plotid, ends_with("v"))
names(rain_v_days) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain_v_days <- rain_v_days %>% gather(year, rain_days_v, -plotid)

rain_r_days <- ppt_days@data %>% select(plotid, ends_with("r"))
names(rain_r_days) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
rain_r_days <- rain_r_days %>% gather(year, rain_days_r, -plotid)

rain_total <- left_join(rain2d_total, 
                        left_join(rain3d_total, rain_v_total, 
                                  by = c("plotid","year")),
                        by = c("plotid","year"))
rain_total <- left_join(rain_total, rain_r_total, by = c("plotid", "year"))
rain_total <- left_join(rain_total, rain_psm_total, by = c("plotid", "year"))
summary(rain_total)# NAs are due to current PRISM data only going through 2010
unique(rain_total$plotid)# 202 plots
unique(rain_total$year)# 10 sampling seasons

rain_days <- left_join(rain2d_days,
                       left_join(rain3d_days, rain_v_days, 
                                 by = c("plotid", "year")),
                       by = c("plotid", "year"))
rain_days <- left_join(rain_days, rain_r_days, by = c("plotid", "year"))
summary(rain_days)
unique(rain_days$plotid)# 202 plots
unique(rain_days$year)# 10 sampling seasons

rm(list = ls(pattern = "^rain_v"))
rm(list = ls(pattern = "^rain2d"))
rm(list = ls(pattern = "^rain3d"))
rm(list = ls(pattern = "^rain_r"))
rm(rain_psm_total)

write.csv(rain_total, "analysis/data/rain_total_long.csv", row.names = F)
write.csv(rain_days, "analysis/data/rain_days_long.csv", row.names = F)
```

## Explore data for completeness and matching
```{r explore plot_env data, message=FALSE}
summary(plots_env)
str(plots_env)
plots_env <- plots_env %>% rename(plotid = PLOT_ID)

write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
# plots_env <- read.csv("analysis/data/env_gis_variables.csv")
```

Examine temperature data, which is in long format and covers sample years 2004 through 2014. This means that I will need to exclude disease-related sampling from 2003 of the ecological data (NAs should fill-in if missing data when joining).
```{r examine temperature & rain data}
# Temperature Data
summary(rs_temps)
summary(summer_temps)

# Rainfall Data
summary(ppt_days)
summary(ppt_total)
rain_df <- merge(ppt_days@data, ppt_total@data)
summary(rain_df)

unique(rs_temps$plotid)# 201 plots
unique(summer_temps$plotid)# 201 plots
unique(rain_df$plotid)# 202 plots

tempsid <- as.character(unique(rs_temps$plotid))
rainid <- as.character(unique(rain_df$plotid))
rainid %in% tempsid
rainid[62]# BUSH01 still in the rain data
rain_df <- filter(rain_df, plotid != "BUSH01")

@@

## Investigating `NAs` in the `plots_temps` data set
A number of observations for above 25 threshold have NAs. I believe that these are in fact truly zero observations. NAs occurred for the 2010, 2011, and 2014 seasons, and the only NAs in the data set are associated with the above 25 threshold. I do believe that it is possible that the temperature was never above 25 C during the wet season November - May in some years.
```{r investigate NAs, eval=FALSE, echo=FALSE}
# summary(plots_env)
# filter(plots_env, plotid == "BUSH01")

# BUSH01 plot included. I think we have limited ecological data for this plot, and no microclimate data. So, I am going to exclude it from all the data sets
# plots_env <- filter(plots_env, plotid != "BUSH01")
stems <- as.tbl(stems)
stems
# stems <- select(stems, -X) 
unique(filter(stems, plot_status == "Abandoned")$plotid)
stems <- stems %>% filter(plotid != "bush01", plotid != "ponti01", 
                                plotid != "sweet01")
stems$plotid <- droplevels(stems$plotid)

rs_temps %>% filter(is.na(hrs_abv25))
# I am replacing the NAs with 0, assuming true observations, i.e. there were zero hours where temperatures were above 25C
rs_temps[is.na(rs_temps)] <- 0
```

### Save data with BUSH01 excluded
```{r save data}
write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
write.csv(rain_df, "analysis/data/rain_data_long.csv", row.names = F)
write.csv(rs_temps, "analysis/data/rs_temps_data.csv", row.names = F)
write.csv(summer_temps, "analysis/data/summer_temps_data.csv", row.names = F)
# write.csv(stems, "analysis/data/stems_noBUSH01.csv", row.names = F)
# rm(plots_temps); rm(plots_rain)
```

## Starting Point with Revised Data: Reload data that has BUSH01 excluded
Still need to load the .RData source at the beginning of this document, line 32.
```{r reload data}
plots_env <- read.csv("analysis/data/env_gis_variables.csv")
rain_df <- read.csv("analysis/data/rain_data_long.csv")
rs_temps <- read.csv("analysis/data/rs_temps_data.csv")
summer_temps <- read.csv("analysis/data/summer_temps_data.csv")
# stems <- read.csv("analysis/data/stems_noBUSH01.csv")
```

@@

I can use the stem-level data to calculate plot level leaf count variables (sum, average) for each plot and year.
```{r calculate plot level leaf counts}
live_umca <- bay_laurel %>% filter(stem_status == "Alive")
filter(live_umca, is.na(umca_slc))
# Spot checking some of these shows that NAs are reasonable
umca_plots <- left_join(
      umca_plots, 
      live_umca %>% 
            group_by(plotid, year) %>% 
            summarise(avg_lfct = mean(umca_slc, na.rm = T),
                      tot_lfct = sum(umca_slc, na.rm = T)),
      by = c("plotid", "year"))
summary(umca_plots)
```

Explore the NAs in the UMCA leaf counts.
```{r examine lfct NAs}
filter(umca_plots, is.na(avg_lfct))
filter(bay_laurel, plotid == "yahng02", stem_status == "Alive")
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
# I'm removing it manually from the `umca_plots` data frame
umca_plots <- umca_plots %>% filter(avg_lfct != "NaN")
summary(umca_plots)
```

@@

# Add rainfall data to UMCA plots data frame
summary(rain_total)
unique(rain_total$plotid)
rain_total$plotid <- tolower(rain_total$plotid)
summary(rain_days)
unique(rain_days$plotid)
rain_days$plotid <- tolower(rain_days$plotid)

rain_total <- rename(rain_total, sample_year = year)
rain_total$sample_year <- as.character(rain_total$sample_year)
rain_days <- rename(rain_days, sample_year = year)
rain_days$sample_year <- as.character(rain_days$sample_year)

@@

filter(umca_plots, is.na(rain_tot_2d))
# Same case of NAs in rainfall due to data from 2003
```

@@

umca_plots <- left_join(umca_plots, os.us.diversity, by = c("plotid"))
summary(umca_plots)
filter(umca_plots, is.na(H.2014))$plotid
## NAs due to plots abandoned/lost between 2005 and 2014; 4 unique plots
filter(umca_plots, is.na(us.H.2005))
## NAs due to plot having '0' for understory richness
filter(umca_plots, is.na(us.H.2011))
## NAs due to 2 plots being abaondoned prior to 2011 & one plot having '0' for understory richness

head(canopy_cover)
canopy_cover <- select(canopy_cover, -X)
canopy_cover <- rename(canopy_cover, sample_year = year)
summary(canopy_cover)
unique(filter(canopy_cover, sample_year == 2006)$plotid)
canopy_cover$sample_year <- as.character(canopy_cover$sample_year)

umca_plots <- left_join(umca_plots, canopy_cover, by = c("plotid", "sample_year"))
summary(umca_plots)# NAs because canopy cover was not measured every year
```

@@

pairs(select(umca_plots, candens15m, pct_cvr, veght15m), 
      lower.panel = panel.cor, diag.panel = panel.hist, 
      main = "Canopy Data - UMCA Plots")
cor(umca_plots$candens15m, umca_plots$pct_cvr, use = "complete.obs")
cor(umca_plots$veght15m, umca_plots$pct_cvr, use = "complete.obs")

pairs(na.omit(select(umca_plots, avg_lfct, tot_lfct, contains("hrs"), contains("rain_tot"), H.2005, H.2014)),
      lower.panel = panel.cor, diag.panel = panel.hist, 
      main = "Leaf Count, Temperature, Rainfall, Diversity - UMCA Plots")
## Rainfall and temperature all have fairly weak correlations (<0.36)
## The diversity metrics are the only correlates with leaf count >0.20
```

@@

# Add bay laurel infection data to oak plots
names(umca_plots)
oak_sod <- left_join(oak_sod, 
                     select(umca_plots, plotid, sample_year, uninfected_bay_ct, infected_bay_ct, tot_bay, avg_lfct, tot_lfct), 
                     by = c("plotid", "sample_year"))
summary(oak_sod) # the NA's for the bay laurel variables are legitimately zeroes b/c there are no bay in some of these plots
filter(oak_sod, is.na(tot_bay), sample_year != 2003)
# Change NAs of bay laurel in oak only plots to zeroes
oak_sod$uninfected_bay_ct[is.na(oak_sod$uninfected_bay_ct)] <- 0
oak_sod$infected_bay_ct[is.na(oak_sod$infected_bay_ct)] <- 0
oak_sod$tot_bay[is.na(oak_sod$tot_bay)] <- 0

@@

os.us.diversity
oak_sod <- left_join(oak_sod, os.us.diversity, by = c("plotid"))
summary(oak_sod)
filter(oak_sod, is.na(avg_tmax_rs))# NAs are plots sampled in 2003
filter(oak_sod, is.na(H.2014))# NAs are plots lost before 2014
summary(filter(oak_sod, is.na(us.J.2011)))# zero or missing data for understory richness
# oak_sod <- filter(oak_sod, plotid != "duer01", plotid != "ponti01", plotid != "sweet01")

@@

filter(oak_sod, is.na(H.2014))$plotid
## plots that were decommissioned prior to 2014
filter(oak_sod, is.na(us.H.2011))$plotid
## plots that have legit 0 for richness or were decommissioned prior to 2011

canopy_cover$sample_year <- as.integer(canopy_cover$sample_year)
oak_sod <- left_join(oak_sod, canopy_cover, by = c("plotid", "sample_year"))

@@

summary(filter(oak_sod, is.na(rain_tot_psm))) # all obs after 2010
summary(filter(oak_sod, is.na(us.J.2011)))
filter(oak_sod, is.na(us.J.2011))$plotid# ann06, mroth03, sumtv02
# ann06 & mroth03 are legitimate zeroes for understory richness in 2011
# sumtv02 was abandoned after 2008, so it should be an NA for later years

@@

### Update path analysis data
Updating the path analysis data with the temperature during wet days variables.
```{r update data}
# load("pathmodel_data_20160112.RData")
wet_hrs <- read.csv("analysis/data/wet-days-temperature-variables.csv")
save.image("~/GitHub/superspreaders/pathmodel_data.RData")
```

>>>>

# Canopy cover from densiometer readings
## I've decided to not use these since we have high-resolution LiDAR data of canopy density
# canopy_cover <- read.csv("analysis/data/plot_canopy_cover_dens.csv")
```
```{r load physical data, echo=FALSE, message=F}
# Physical environment: topography
library(rgdal); library(plyr); library(dplyr)
plots_shp <- readOGR("analysis/data/plot_gis/", "soco_plots_metric")
plots_env <- plots_shp@data
summary(plots_env)
plots_env <- select(plots_env, -cat, -X, -Y)

# Elevation and vegetation height are in feet, slope is degrees, and canopy density is the proportion of the cell covered by canopy (0-1)
plots_topo.shp <- readOGR("analysis/data/plot_gis/", "plots202_vegtopo")
plots_vegtopo <- plots_topo.shp@data
summary(plots_vegtopo)
plots_vegtopo <- select(plots_vegtopo, -cat, -flowacc15m, -facc_ws15m, -slope15m) %>% rename(lon_utm = X, lat_utm = Y)
```
```{r load weather data, echo=FALSE}
# Temperature: rainy season hour summaries (long format for each year)
rs_temps <- read.csv("analysis/data/rs_select_temps.csv")
summary(rs_temps)
# rs_temps <- select(rs_temps, -X)

summer_temps <- read.csv("analysis/data/summer_select_temps.csv")
summary(summer_temps)
# summer_temps <- select(summer_temps, -X)

wet_hrs <- read.csv("analysis/data/wet-days-temperature-variables.csv")

# Rainfall: interpolated/estimated rainfall for each season
load("~/Documents/soco_ppt/plot_ppt_estimates.RData")
rm(ppt_days)
rm(ppt_total)
summary(rainfall)# use newly calculated values
rainfall$plotid <- tolower(rainfall$plotid)
# plots_rain <- read.csv("analysis/data/plot202rain.csv")
# summary(plots_rain)
# plots_rain <- select(plots_rain, -cat)
# pairs(select(plots_rain, -PLOT_ID), 
#       lower.panel = panel.cor, 
#       diag.panel = panel.hist)
```

## Join remotely sensed and derived physical data
LiDAR products, variables derived from DEMs
```{r join remotely sensed data}
plots_env <- inner_join(plots_env, plots_vegtopo, by = "PLOT_ID")
summary(plots_env)
plots_env$elev_m <- plots_env$elev15m * .3048
# plots_env <- select(plots_env, -elevation, -tmi) #remove elevation, tmi from old derivation (I think a legacy 10m USGS DEM)
plots_env$veght_m <- plots_env$veght15m * .3048
plots_env <- select(plots_env, -elevation, -elev15m, -veght15m, -tmi) %>% 
      rename(plotid = PLOT_ID)
plots_env$plotid <- tolower(plots_env$plotid)
rm(plots_vegtopo)
```

## Convert rain data from wide to long
```{r rain wide to long}
library(tidyr)
rain2d_rs_total <- rainfall@data %>% 
      select(plotid, contains("rs"),  -ends_with("v"))
names(rain2d_rs_total) <- c("plotid",paste(c(2004:2014), sep = ","))
rain2d_rs_total <- rain2d_rs_total %>% gather(year, rain2d_total, -plotid)

rain2d_ss_total <- rainfall@data %>% 
      select(plotid, contains("ss"),  -ends_with("v"))
names(rain2d_ss_total) <- c("plotid",paste(c(2004:2014), sep = ","))
rain2d_ss_total <- rain2d_ss_total %>% gather(year, rain2d_total, -plotid)

# rain3d_total <- ppt_total@data %>% select(plotid, ends_with("3d"))
# names(rain3d_total) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
# rain3d_total <- rain3d_total %>% gather(year, rain_tot_3d, -plotid)

rain_rs_v_total <- rainfall@data %>% 
      select(plotid, ends_with("v")) %>% 
      select(plotid, contains("rs")) %>%
      select(plotid, starts_with("rain"))
names(rain_rs_v_total) <- c("plotid",paste(c(2004:2014), sep = ","))
rain_rs_v_total <- rain_rs_v_total %>% gather(year, rain_v_total, -plotid)

rain_ss_v_total <- rainfall@data %>% 
      select(plotid, ends_with("v")) %>% 
      select(plotid, contains("ss")) %>%
      select(plotid, starts_with("rain"))
names(rain_ss_v_total) <- c("plotid",paste(c(2004:2014), sep = ","))
rain_ss_v_total <- rain_ss_v_total %>% gather(year, rain_v_total, -plotid)

# rain_r_total <- ppt_total@data %>% select(plotid, ends_with("r"))
# names(rain_r_total) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
# rain_r_total <- rain_r_total %>% gather(year, rain_tot_r, -plotid)

# rain_psm_total <- ppt_total@data %>% select(plotid, contains("sum"))
# names(rain_psm_total) <- c("plotid", paste(c(2004:2010), sep = ","))
# rain_psm_total <- rain_psm_total %>% gather(year, rain_tot_psm, -plotid)

# rain2d_days <- ppt_days@data %>% select(plotid, ends_with("2d"))
# names(rain2d_days) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
# rain2d_days <- rain2d_days %>% gather(year, rain_days_2d, -plotid)

# rain3d_days <- ppt_days@data %>% select(plotid, ends_with("3d"))
# names(rain3d_days) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
# rain3d_days <- rain3d_days %>% gather(year, rain_days_3d, -plotid)

rain_rs_v_days <- rainfall@data %>% 
      select(plotid, contains("dys")) %>% 
      select(plotid, ends_with("v")) %>% 
      select(plotid, contains("rs"))
names(rain_rs_v_days) <- c("plotid",paste(c(2004:2014), sep = ","))
rain_rs_v_days <- rain_rs_v_days %>% gather(year, rain_days_v, -plotid)

rain_ss_v_days <- rainfall@data %>% 
      select(plotid, contains("dys")) %>% 
      select(plotid, ends_with("v")) %>% 
      select(plotid, contains("ss"))
names(rain_ss_v_days) <- c("plotid",paste(c(2004:2014), sep = ","))
rain_ss_v_days <- rain_ss_v_days %>% gather(year, rain_days_v, -plotid)

# rain_r_days <- ppt_days@data %>% select(plotid, ends_with("r"))
# names(rain_r_days) <- c("plotid",paste(c(2004:2012,2014), sep = ","))
# rain_r_days <- rain_r_days %>% gather(year, rain_days_r, -plotid)

rain_rs_total <- left_join(rain2d_rs_total,
                           rain_rs_v_total,
                           by = c("plotid","year"))

rain_ss_total <- left_join(rain2d_ss_total,
                           rain_ss_v_total,
                           by = c("plotid","year"))

# rain_total <- left_join(rain_total, rain_r_total, by = c("plotid", "year"))
# rain_total <- left_join(rain_total, rain_psm_total, by = c("plotid", "year"))
# summary(rain_total)# NAs are due to current PRISM data only going through 2010
# unique(rain_total$plotid)# 202 plots
# unique(rain_total$year)# 10 sampling seasons

# rain_days <- left_join(rain2d_days,
#                        left_join(rain3d_days, rain_v_days, 
#                                  by = c("plotid", "year")),
#                        by = c("plotid", "year"))
# rain_days <- left_join(rain_days, rain_r_days, by = c("plotid", "year"))
# summary(rain_days)
# unique(rain_days$plotid)# 202 plots
# unique(rain_days$year)# 10 sampling seasons

# rm(list = ls(pattern = "^rain_v"))
# rm(list = ls(pattern = "^rain2d"))
# rm(list = ls(pattern = "^rain3d"))
# rm(list = ls(pattern = "^rain_r"))
# rm(rain_psm_total)

write.csv(rain_rs_total, "analysis/data/rainy_season_rainfall_total_long.csv", row.names = F)
write.csv(rain_ss_total, "analysis/data/sampling_season_rainfall_total_long.csv", row.names = F)
write.csv(rain_rs_v_days, "analysis/data/rainy_season_wet_days_long.csv", row.names = F)
write.csv(rain_ss_v_days, "analysis/data/sampling_season_wet_days_long.csv", row.names = F)
```

## Explore data for completeness and matching
```{r explore plot_env data, message=FALSE}
summary(plots_env)
str(plots_env)
# plots_env <- plots_env %>% rename(plotid = PLOT_ID)

# write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
# plots_env <- read.csv("analysis/data/env_gis_variables.csv")
```

Examine temperature data, which is in long format and covers sample years 2004 through 2014. This means that I will need to exclude disease-related sampling from 2003 of the ecological data (NAs should fill-in if missing data when joining).
```{r examine temperature & rain data}
# Temperature Data
summary(rs_temps)
rs_temps$plotid <- tolower(rs_temps$plotid)
summary(summer_temps)
summer_temps$plotid <- tolower(summer_temps$plotid)

# Rainfall Data
rainy_season_ppt <- merge(rain_rs_total, rain_rs_v_days)
summary(rainy_season_ppt)

length(unique(rs_temps$plotid))# 201 plots
length(unique(summer_temps$plotid))# 201 plots
length(unique(rainy_season_ppt$plotid))# 202 plots

tempsid <- tolower(as.character(unique(rs_temps$plotid)))
rainid <- as.character(unique(rainy_season_ppt$plotid))
rainid %in% tempsid
rainid[62]# BUSH01 still in the rain data
rainy_season_ppt <- filter(rainy_season_ppt, plotid != "bush01")

@@

## Investigating `NAs` in the temperature data set
A number of observations for above 25 threshold have NAs. I believe that these are in fact truly zero observations. NAs occurred for the 2010, 2011, and 2014 seasons, and the only NAs in the data set are associated with the above 25 threshold. I do believe that it is possible that the temperature was never above 25 C during the wet season November - May in some years.
```{r investigate NAs, eval=FALSE, echo=FALSE}
rs_temps %>% filter(is.na(hrs_abv25))
# I am replacing the NAs with 0, assuming true observations, i.e. there were zero hours where temperatures were above 25C
rs_temps[is.na(rs_temps)] <- 0
```

### Save data with BUSH01 excluded
```{r save data}
stems <- as.tbl(stems)
stems
# stems <- select(stems, -X) 
# unique(filter(stems, plot_status == "Abandoned")$plotid)
stems <- stems %>% filter(plotid != "bush01", plotid != "ponti01", 
                                plotid != "sweet01")
stems$plotid <- droplevels(stems$plotid)
# write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
# write.csv(rainy_season_ppt, "analysis/data/rain_data_long.csv", row.names = F)
# write.csv(rs_temps, "analysis/data/rs_temps_data.csv", row.names = F)
# write.csv(summer_temps, "analysis/data/summer_temps_data.csv", row.names = F)
# write.csv(stems, "analysis/data/stems_noBUSH01.csv", row.names = F)
# rm(plots_temps); rm(plots_rain)
```

## Starting Point with Revised Data: Reload data that has BUSH01 excluded
Still need to load the .RData source at the beginning of this document, line 32.
```{r reload data}
# plots_env <- read.csv("analysis/data/env_gis_variables.csv")
# rain_df <- read.csv("analysis/data/rain_data_long.csv")
# rs_temps <- read.csv("analysis/data/rs_temps_data.csv")
# summer_temps <- read.csv("analysis/data/summer_temps_data.csv")
# stems <- read.csv("analysis/data/stems_noBUSH01.csv")
```

@@

I can use the stem-level data to calculate plot level leaf count variables (sum, average) for each plot and year.
```{r calculate plot level leaf counts}
live_umca <- bay_laurel %>% filter(stem_status == "Alive")
filter(live_umca, is.na(umca_slc))
# Spot checking some of these shows that NAs are due to missing data, i.e. they should not be changed to zeroes
umca_plots <- left_join(
      umca_plots, 
      live_umca %>% 
            group_by(plotid, year) %>% 
            summarise(avg_lfct = mean(umca_slc, na.rm = T),
                      tot_lfct = sum(umca_slc, na.rm = T)),
      by = c("plotid", "year"))
summary(umca_plots)
```

Explore the NAs in the UMCA leaf counts.
```{r examine lfct NAs}
filter(umca_plots, is.na(avg_lfct))
filter(bay_laurel, plotid == "yahng02", stem_status == "Alive")
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
# I'm removing it manually from the `umca_plots` data frame
umca_plots <- umca_plots %>% filter(avg_lfct != "NaN")
# filter(umca_plots, plotid == "yahng02")
summary(umca_plots)
```

@@

# Add rainfall data to UMCA plots data frame
summary(rain_rs_total)
summary(rain_ss_total)
# unique(rain_rs_total$plotid)
# unique(rain_ss_total$plotid)

summary(rain_rs_v_days)
summary(rain_ss_v_days)
# unique(rain_rs_v_days$plotid)
# unique(rain_ss_v_days$plotid)

rain_rs_total <- rename(rain_rs_total, sample_year = year,
                        rs2d_total = rain2d_total,
                        rs_v_total = rain_v_total)
rain_ss_total <- rename(rain_ss_total, sample_year = year,
                        ss2d_total = rain2d_total,
                        ss_v_total = rain_v_total)
rain_total <- left_join(rain_rs_total, rain_ss_total, by = c("plotid", "sample_year"))
# rain_total$sample_year <- as.character(rain_total$sample_year)
rain_rs_v_days <- rename(rain_rs_v_days, sample_year = year, rs_days = rain_days_v)
rain_ss_v_days <- rename(rain_ss_v_days, sample_year = year, ss_days = rain_days_v)
rain_days <- left_join(rain_rs_v_days, rain_ss_v_days, by = c("plotid", "sample_year"))
# rain_days$sample_year <- as.character(rain_days$sample_year)

@@

filter(umca_plots, is.na(rs2d_total))
# Same case of NAs in rainfall due to data sampled in 2003
```

@@

umca_plots <- left_join(umca_plots, os.us.diversity, by = c("plotid"))
summary(umca_plots)
unique(filter(umca_plots, is.na(H.2014))$plotid)
## NAs due to plots abandoned/lost between 2005 and 2014; 4 unique plots
filter(umca_plots, is.na(us.H.2005))
## NAs due to plot having '0' for understory richness
filter(umca_plots, is.na(us.H.2011))
## NAs due to 2 plots being abaondoned prior to 2011 & one plot having '0' for understory richness

# head(canopy_cover)
# canopy_cover <- select(canopy_cover, -X)
# canopy_cover <- rename(canopy_cover, sample_year = year)
# summary(canopy_cover)
# unique(filter(canopy_cover, sample_year == 2006)$plotid)
# canopy_cover$sample_year <- as.character(canopy_cover$sample_year)
# 
# umca_plots <- left_join(umca_plots, canopy_cover, by = c("plotid", "sample_year"))
# summary(umca_plots)# NAs because canopy cover was not measured every year
```

@@

pairs(select(umca_plots, candens15m, veght_m), 
      lower.panel = panel.cor, diag.panel = panel.hist, 
      main = "Canopy Data - UMCA Plots")
# cor(umca_plots$candens15m, umca_plots$pct_cvr, use = "complete.obs")
# cor(umca_plots$veght15m, umca_plots$pct_cvr, use = "complete.obs")

pairs(na.omit(select(umca_plots, avg_lfct, tot_lfct, contains("hrs"), contains("_total"), H.2005, H.2014)),
      lower.panel = panel.cor, diag.panel = panel.hist, 
      main = "Leaf Count, Temperature, Rainfall, Diversity - UMCA Plots")
plot(umca_plots$hrs_abv25ds, log1p(umca_plots$tot_lfct))
cor(umca_plots$hrs_abv25ds, log1p(umca_plots$tot_lfct), use = "complete")
```

@@

# Add bay laurel infection data to oak plots
names(umca_plots)
oak_sod <- left_join(oak_sod, 
                     select(umca_plots, plotid, sample_year, uninfected_bay_ct, infected_bay_ct, tot_bay, tot_lfct), 
                     by = c("plotid", "sample_year"))
summary(oak_sod) # the NA's for the bay laurel variables are legitimately zeroes b/c there are no bay in some of these plots
filter(oak_sod, is.na(tot_bay), sample_year != 2003)
# Change NAs of bay laurel in oak only plots to zeroes
oak_sod$uninfected_bay_ct[is.na(oak_sod$uninfected_bay_ct)] <- 0
oak_sod$infected_bay_ct[is.na(oak_sod$infected_bay_ct)] <- 0
oak_sod$tot_bay[is.na(oak_sod$tot_bay)] <- 0

@@

os.us.diversity
oak_sod <- left_join(oak_sod, os.us.diversity, by = c("plotid"))
summary(oak_sod)
summary(filter(oak_sod, is.na(avg_tmax_rs)))# NAs are plots sampled in 2003
unique(filter(oak_sod, is.na(H.2014))$plotid)# NAs are plots lost before 2014
summary(filter(oak_sod, is.na(us.J.2011)))# zero or missing data for understory richness
# oak_sod <- filter(oak_sod, plotid != "duer01", plotid != "ponti01", plotid != "sweet01")

@@

unique(filter(oak_sod, is.na(H.2014))$plotid)
## plots that were decommissioned prior to 2014 sampling
filter(oak_sod, is.na(us.H.2011))$plotid
## plots that have legit 0 for understory richness or were decommissioned prior to 2011

# canopy_cover$sample_year <- as.integer(canopy_cover$sample_year)
# oak_sod <- left_join(oak_sod, canopy_cover, by = c("plotid", "sample_year"))

@@

# summary(filter(oak_sod, is.na(rain_tot_psm))) # all obs after 2010
# summary(filter(oak_sod, is.na(us.J.2011)))
unique(filter(oak_sod, is.na(us.J.2011))$plotid)# ann06, mroth03, sumtv02
# ann06 & mroth03 are legitimate zeroes for understory richness in 2011
# i must have excluded mcneil03 from the diversity calculations
# sumtv02 was abandoned after 2008, so it should be an NA for later years

@@

### Update path analysis .Rdata object
Updating the path analysis data with the temperature during wet days variables.
```{r update data}
# load("~/GitHub/superspreaders/pathmodel_data.RData")

# Add untagged dbh data
untagged_stems <- read.csv("analysis/data/untagged_stems_corrected.csv")
untagged_dbh_2005 <- read.csv("analysis/data/untagged_dbh2005_summary.csv")
untagged_dbh_2014 <- read.csv("analysis/data/untagged_dbh2014_summary.csv")

# remove unnecessary data
rm(list = ls(pattern = "^rain_rs"))
rm(list = ls(pattern = "^rain_ss"))
rm(list = ls(pattern = "^rain2d"))
rm(list = ls(pattern = "*ds_lag"))
rm(list = ls(pattern = "*0_rs_lag")); rm(avgtmax_rs_lag)
rm(list = ls(pattern = "*wet_lag"))

save.image("~/GitHub/superspreaders/pathmodel_data.RData")
```
